Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-18 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, some of which have been developed recently~\cite{Quaife2018} and others that await resolution. {\bf An ultimate goal of this proposal is to achieve high-fidelity simulations of $\boldsymbol{\mathcal{O}(100-1000)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.}

	There has been a recent spark in interest in dynamic erosion of porous media, both theoretically and experimentally. In fact, experiments on erosion of colloidal systems are currently being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments are closely related to our proposed computational study, thus offering a great opportunity for comparison. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute (see Section~\ref{sec:experiments}), which would help to guide, calibrate, and validate the computations (see letter of collaboration).

@@

%Very recently, a second paper has been submitted to the Journal of Fluid Mechanics~\cite{chi-moo-qua2019}, where additional numerical methods are developed and characterizations of transport are presented.

@@

%----------------------------------------------------------------------%
\subsection{Mechanical and Chemical Erosion of a Porous Medium}
With the development of efficient and high-fidelity numerical methods to solve the Stokes equations, we are able to simulate erosion. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{C}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques.  The boundary integral solver is accelerated by the fast multipole method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

@@

Figs.~\ref{80circ8vort} and~\ref{80circ8vel} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables closer contact between bodies and thus more bodies. The color in Fig~\ref{80circ8vort} shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity (Fig.~\ref{80circ8vel}).  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vel.pdf}
\caption{\label{80circ8vel} High-fidelity numerical simulations of a configuration of densely-packed solid bodies undergoing mechanical erosion due to an intervening Stokes flow. The color indicates velocity magnitude which highlights the channel formation process. The solid bodies develop flat faces and sharp corners during evolution. }
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

An immediate goal of the proposed work is to extend the range of physical effects captured in these simulations. A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. However, the implementation of the quadrature method does not fully use FMM acceleration. The numerical challenge is developing the required FMM expansions for the velocity, vorticity, and deformation tensor layer potentials. Since we require several layer potentials with different kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and its HPC implementations that are readily available (for example, see \url{http://pvfmm.org}). Other challenges that will be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to allow for stable simulations to user-specified tolerances. PI Quaife has experience developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.

>>>>

\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, some of which have been developed recently~\cite{Quaife2018} and others that await resolution.

{\bf An ultimate goal of this proposal is to achieve high-fidelity simulations of $\boldsymbol{\mathcal{O}(100-1000)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.} A major step towards this goal is illustrated in Fig.~\ref{80circ8vel}, which features high-fidelity erosion simulations of an initial configuration of 80 bodies with a volume fraction of $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities, for example contaminants.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vel.pdf}
\caption{\label{80circ8vel} High-fidelity numerical simulations of a configuration of densely-packed solid bodies undergoing mechanical erosion due to an intervening Stokes flow. The color indicates velocity magnitude which highlights the channel formation process. The solid bodies develop flat faces and sharp corners during evolution. }
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

 \todo[inline]{I guess I really took to your strategy of including an example simulation early on in the introduction to draw in the readers interest. So I moved this figure up to the introduction and added some provocative text. Plus there are some visual similarities with the geophysical images in Fig. 1, so it is nice to see them close to one another.}

	In addition to the work of the PIs, there has been a recently sparked interest in dynamic erosion of porous media, as experiments on erosion of colloidal systems are currently being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments are closely related to our proposed computational study, thus offering a great opportunity for comparison. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute (see Section~\ref{sec:experiments}), which would help to guide, calibrate, and validate the computations (see letter of collaboration).

@@

@@

%----------------------------------------------------------------------%
\subsection{Mechanical and Chemical Erosion of a Porous Medium}

The development of accurate and efficient numerical methods to solve the Stokes equations enables high-fidelity simulations of the erosion process. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{C}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques.  The boundary integral solver is accelerated by the fast multipole method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

@@

Figs.~\ref{80circ8vort} and~\ref{80circ8vel} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables closer contact between bodies and thus more bodies. The color in Fig~\ref{80circ8vort} shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity (Fig.~\ref{80circ8vel}).  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is made even more apparent by viewing the velocity magnitude as in Fig.~\ref{80circ8vel}.

@@

{An immediate goal of the proposed work is to extend the range of physical effects captured in these simulations.} A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. However, the implementation of the quadrature method does not fully use FMM acceleration. The numerical challenge is developing the required FMM expansions for the velocity, vorticity, and deformation tensor layer potentials. Since we require several layer potentials with different kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and its HPC implementations that are readily available (for example, see \url{http://pvfmm.org}). Other challenges that will be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to allow for stable simulations to user-specified tolerances. PI Quaife has experience developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.
